Really Good Title

### 2019-01-07 documentation/paper.tex
minor workwork
Holy hell it's late

\documentclass[11pt,a4paper]{scrartcl}
%{{{ general stuff
\usepackage[a4paper,bindingoffset=0.2in,%
            left=1in,right=1in,top=1in,bottom=1in,%
            footskip=.25in]{geometry}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\usepackage{float} % use H in figure placement
\pagenumbering{gobble}

@@

\date{}
\title{todo: think about good title}
\author{Felix Bartel and D\'avid Kerekes}

@@

\maketitle

@@

Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years. It's provable quality of being an universal function approximator makes it a sensible choice for various applications from control to . The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{some_papers_about_backprop}, and there are various attempts at improving or replacing it \cite{that_paper_where_ga_beats_backprop}.
\todo{Add a sentence of context for \cite{Sims94evolving3d} and \cite{Garis90geneticprogramming}}

In this work we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
A big advantage of this scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
This can for instance be given by the outcome of a game.
Our game of choice is Blobby Volley 2 which is an open source, simple two-dimensional volleyball game written in C++ with lua support. See Figure \ref{fig:screenshot}.

@@

\section{Implementation}
The genotype-representation of the network is accomplished through python objects. This allows a straightforward implementation of a wide variety of genetic operators. Since our algorithm is constrained overwhelmingly by evaluation time, these inefficient but easy to modify structures do not impact performance.
As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.

\section{Standard genetic algorithm}

@@

\subsection{Scheme}
The algorithm worked with a population size of 100, with 150 offspring after crossover. Our survivor selection strategy was elitism, selecting from the pool of children, plus the 10 best parents from the last generation.

@@

\subsection{Mutation}
Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases. \todo{Default values, rate of change}

\subsection{Crossover}
As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection. \todo{cite roulett wheel sel.}

\begin{figure}[H]
\center
\begin{tabular}{ccc}
\texttt{reduced} & \texttt{com\_11} & \texttt{gintonicV9} \\
\includegraphics[width=0.3\textwidth]{img/standard_reduced.png} &
\includegraphics[width=0.3\textwidth]{img/standard_com_11.png} &
\includegraphics[width=0.3\textwidth]{img/standard_gintonicV9.png} \\
\texttt{hyp014} & \texttt{Union} & \texttt{trainer} \\
\includegraphics[width=0.3\textwidth]{img/standard_hyp014.png} &
\includegraphics[width=0.3\textwidth]{img/standard_Union.png} &
\includegraphics[width=0.3\textwidth]{img/standard_trainer.png}
\end{tabular}
\caption{Fitness density with indicated minimum, average and maximum while training against different bots. \todo{is this true?}}
\label{fig:standard}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/standard.png}
\caption{trained bots against all other \todo{explain this pls}}
\label{fig:standard}
\end{figure}

\todo{coclusion: bots too hard we therefore reward gameplay by the *\_5t bots}

\todo{introduce the goal of an overall good bot and explain the plot along the way (added fitnesses (score of 12 would mean winneing every point against every bot))}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/standard_5t.png}
\caption{Trained bots against all other}
\label{fig:standard_5t_aa}
\end{figure}

\todo{conclusion: better, and not only for 5t bots (algorithm has something to grab on)}

\todo{overall: not very pretty graphs -> need self adaption}

\section{Self adapt 1}

\todo{explain what we are doing}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_1.png}
\caption{trained bots against all other}
\label{fig:self_adapt_1_aa}
\end{figure}

\todo{we can see that the algorithm is converging}
\todo{because trainer is beatable, the nn has something to grab on and performs way better thatn the others}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
\caption{trained bots against all other}
\label{fig:self_adapt_2_5t_aa}
\end{figure}

\todo{now it's pretty}
\todo{trainer\_5t bad because it is very easy to beat and therefore we can get loast in a local optimum real quick}

\section{Self adapt 2}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2.png}
\caption{trained bots against all other}
\label{fig:self_adapt_2_aa}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
\caption{trained bots against all other}
\label{fig:self_adapt_2_5t_aa}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/change.png}
\caption{trained bots against all other}
\label{fig:change}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.3\textwidth]{img/self_training_1.png}
\caption{trained bots against all other}
\label{fig:self_training_1}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_training_2.png}
\caption{trained bots against all other}
\label{fig:self_training_2}
\end{figure}

@@

>>>>

\documentclass[11pt,a4paper]{scrartcl}
%{{{ general stuff
\usepackage[a4paper,bindingoffset=0.2in,%
            left=1in,right=1in,top=1in,bottom=1in,%
            footskip=.25in]{geometry}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\usepackage{float} % use H in figure placement
% \pagenumbering{gobble}

@@

\date{\today}
\title{Really Good Title}
\subtitle{Genetic Algorithms and Evolutionary Computing Project}
\author{Felix Bartel and D\'avid Kerekes}

@@

\maketitle
\todo{think about good title}

@@

Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years. It's provable quality of being an universal function approximator makes it a sensible choice for various applications. The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{some_papers_about_backprop}, and there are various attempts at improving or replacing it \cite{that_paper_where_ga_beats_backprop}.
\todo{Add a sentence of context for \cite{Sims94evolving3d} and \cite{Garis90geneticprogramming}}

In this project we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
A big advantage of this scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
This can for instance be given by the outcome of a game.
Our game of choice is Blobby Volley 2 which is an open source, simple two-dimensional volleyball game written in C++ with lua support. See Figure \ref{fig:screenshot}.

@@

\section{Implementation}
The genotype-representation of the network is accomplished through python objects. This allows a straightforward implementation of a wide variety of genetic operators. Since our algorithm is constrained overwhelmingly by evaluation time, these inefficient but easy to modify structures do not impact performance.
As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal. In order to achieve this, we modified the original game by replacing the graphics and sound modules with dummies.

\section{Initial genetic algorithm}
Here we describe our starting point for the genetic algorithm. All subsequent versions are based off of this variant, with not mentioned details staying the same.

@@

\subsection{Scheme}
The algorithm worked with a population size of 100, with 150 offspring after crossover. Our survivor selection strategy was elitism, selecting from the pool of children plus the 10 best parents from the last generation.

@@

\subsection{Mutation}
Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases. We selected an individual for mutation with a probability of $0.7$, which then had a per-neuron mutation rate of $0.5$, and a sigma of $0.01$.

\subsection{Crossover}
As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.

\subsection{Numerical tests}
We trained six networks, each against a different bot. We test the generalization of both the training method and the best specimens by evaluating all the networks against all the bots. A score for each network is calculated by summing up it's fitness value for all cases (see Fig. \ref{fig:standard_c}), so a network consistently beating all opponents would receive a score of 12. We expect that a good training method would produce networks that all beat their respective trainers and also fare well against other opponents. The chances of happening upon a network that can exploit a trainer by "accident" (but is susceptible to abuse from a human player for example) decrease exponentially by evaluating on multiple opponents.

\begin{figure}[H]
\center
\begin{tabular}{ccc}
\texttt{reduced} & \texttt{com\_11} & \texttt{gintonicV9} \\
\includegraphics[width=0.3\textwidth]{img/standard_reduced.png} &
\includegraphics[width=0.3\textwidth]{img/standard_com_11.png} &
\includegraphics[width=0.3\textwidth]{img/standard_gintonicV9.png} \\
\texttt{hyp014} & \texttt{Union} & \texttt{trainer} \\
\includegraphics[width=0.3\textwidth]{img/standard_hyp014.png} &
\includegraphics[width=0.3\textwidth]{img/standard_Union.png} &
\includegraphics[width=0.3\textwidth]{img/standard_trainer.png}
\end{tabular}
\caption{Fitness density with indicated minimum, average and maximum while training against different bots.}
\label{fig:standard}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/standard.png}
\caption{Cumulative sum of the trained neural networks against all the bots.}
\label{fig:standard_c}
\end{figure}

We can see that the algorithm achieves reasonable scores against the two easiest opponents, \texttt{trainer} and \texttt{hyp014}. Since we were not able to get significantly better results by hyperparameter-tweaking, we theorized that the discrepancy might come from the trainers being too hard to beat by any initial network mutation, therefore providing the same zero fitness for any single useful mutation. Therefore we introduced a weaker, "benevolent" set of trainers that give up after scoring five points by not returning the ball anymore. This lets a specimen that can pass initial serves back receive a positive fitness score. These bots are marked by appending the \texttt{\_5t} suffix to their name.

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/standard_5t.png}
\caption{Trained networks against all other bots.}
\label{fig:standard_5t_aa}
\end{figure}

As Fig. \ref{fig:standard_5t} and \ref{fig:standard_5t_aa} show, the networks evolved against the benevolent trainers give significantly better performance against the whole batch of bots, reaching twice the maximum cumulative score.

\section{Per-individual self adaptation}

\todo{overall: not very pretty graphs <<-- Why do we decided to do self adaptation?}

We first implemented a method similar to (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value. These were mutated when the individual was selected for mutation. Their initialization was also random, in the range of [0.25, 0.75] for the mutation rate and [0, 0.03] for the sigma values.

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_1.png}
\caption{Trained networks against all other bots.}
\label{fig:self_adapt_1_aa}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
\caption{Trained networks against all other bots.}
\label{fig:self_adapt_1_5t_aa}
\end{figure}

We can see in figures \ref{self_adapt_1_5t} and \ref{self_adapt_1_5t_aa} that the results improve significantly in both cases. Self adaptation causes the networks trained on \texttt{_5t} opponents to get consistent results across the board.

\todo{trainer\_5t does not coincide on the two graphs, what happened? We get a good result against trainer_5t according to the bottom graph}

\section{Per-parameter self adaptation}

We also explored a method similar to (5) in \cite{self_adapt}, where each parameter gets it's own mutation sigma parameter. This produces similar results to the harsh teachers with per-individual self adaptation

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2.png}
\caption{Trained networks against all other bots.}
\label{fig:self_adapt_2_aa}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
\caption{Trained networks against all other bots.}
\label{fig:self_adapt_2_5t_aa}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/change.png}
\caption{Trained networks against all other bots.}
\label{fig:change}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.3\textwidth]{img/self_training_1.png}
\caption{Trained networks against all other bots.}
\label{fig:self_training_1}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[width=0.4\textwidth]{img/self_training_2.png}
\caption{Trained networks against all other bots.}
\label{fig:self_training_2}
\end{figure}

@@

\todo{rationalize not messing around with more mutation and crossover types as we promised}
